\textbf{Beasts of Prey or Rational Animals?\\ Private Governance in Brazil's \emph{Jogo do Bicho}}

### 2018-09-20 paper-bicho/bicho.tex
changes sec1

In this article I offer a rational choice interpretation of the \emph{jogo do bicho}. More specifically, I discuss how \emph{bicheiros} adopt rational strategies to solve collective action problems and maximise their political leverage. I use the case of Rio de Janeiro to illustrate the lottery's profit abilities and its effect in political corruption. The lottery financiers have long been major sponsors of Rio de Janeiro's world-famous Carnival Parade, and several of the country's most high-profile \emph{jogo do bicho} entrepreneurs hail from that state. Furthermore, in no other part of Brazil the patron-client relationship between \emph{bicheiros} and politicians has been so widely publicised.

 to co-opt local politicians attracted by the electoral gains offered by the festival \citep{cavalcanti2006carnaval, queiroz1992carnaval}.

@@

\section{\emph{Jogo do Bicho} as an Emergent Institution}
\label{sec:institution}

\subsection{Historical Background}
\label{sub:background}

The early history of the \emph{jogo do bicho} is a textbook example of spontaneous order. Spontaneous orders are emergent macro-level phenomena that result from voluntary actions of purposive, self-interested individuals utilising their contextual knowledge \citep{boettke1990theory, boettke2005methodological, hayek1945use, hayek1960constitution, hayek1973law, leeson2008coordination, menger1871grundsatze, polanyi1948planning, polanyi1951logic}. Drummond, the Zoological Garden's original owner, designed the basic framework for the \emph{jogo do bicho}; but independent bookmakers were the ones who popularised the game \citep[77]{magalhaes2005ganhou}. Ticket sellers could quickly respond to market signals and then allocate their products where they were more valuable because of the lack of central coordination. Moreover, competition among sellers fostered innovation, and the \emph{bicheiros} invented new game rules to make the lottery more appealing to their customers \citep[61]{mello1989historia}. In this sense, the animal game is the materialisation of an evolutionary process of entrepreneurial discovery in which the interactions that provided the highest value to consumers were preserved over time \citep{boettke2008gordon, boettke2014entrepreneurship, buchanan1964should, hayek1978competition, kirzner1997entrepreneurial}.

However, the \emph{jogo do bicho} only emerged because of historically contingent circumstances. The late nineteenth-century Brazil had four characteristics that explain how the animal game came to being: 1) a growing urban population excluded from the formal labour market; 2) an inflow of immigrants whose extended family networks helped them engage in trade; 3) an expansion of the monetary supply in the first years of the republic (1880s--1890s); and 4) a judicial system that, albeit repressive, had only imperfect law enforcement. Figure \ref{fig:dag} presents a simple directed acyclic graph (DAG) \citep{pearl2009causality} that shows the relationships between these explanatory variables and the development of the \emph{jogo do bicho}.\footnote{The main purpose of direct acyclic graphs is to graphically display the possible links among the exposure variables, confounders, and outcomes \citep{morgan2014counterfactuals, pearl2009causality}. DAGs are transparent by definition, as all theoretical choices made by the researcher are stated explicitly in the model. Each single-headed arrow in a DAG indicates that the variable at the origin causes the variable at the end of the directed edge. Dashed edges suggest that two variables are jointly dependent on unobserved common causes. There are no assumptions regarding the functional form of the relationships, and unless mentioned otherwise, the arrows represent fully non-parametric associations. Variables between two nodes are mediators, and variables pointed at by two or more factors have multiple causes. They are called \emph{colliders}. For the sake of clarity, errors are assumed independent and often excluded from the graphs. For an accessible introduction to DAGs see \citet[chap. 3--4]{morgan2014counterfactuals} and \citet{pearl2016causal}.}

I start with the impact of urban poverty on the animal game. Brazil abolished slavery in the late 1880s, a period in which the country was rapidly urbanising and freed slaves migrated to its growing cities \citep{andrews1991blacks, fausto2014concise, naro1992revision, skidmore1993black}. The former slaves were joined by increasing numbers of Asian and European immigrants \citep{hall1969origins, lesser2013immigration, smith1979ethnic}. Nevertheless, the job market tightened considerably after the \emph{Encilhamento} financial crisis of 1891 \citep{topik2014political, triner2005baring}. During the economic downturn, the informal economy was an obvious destination for the urban poor. Given its widespread popularity, the \emph{jogo do bicho} attracted hopeful entrepreneurs, either Brazilian or foreign-born, who could not enter the formal labour force.\footnote{The underground economy was also more democratic than the formal sector. As \citet[115]{chazkel2011laws} observes, one of the few professions open to poor women and foreigners in the early 1900s was that of street vendor. These vendors used to sell different types of merchandise and many of them would later offer \emph{jogo do bicho} tickets.}

The immigration also influenced the \emph{jogo do bicho} via social ties. Most foreigners who moved to Brazil came from countries, such as Portugal, Spain or Italy, where extended families were the basic form of social organisation \citep{klein1994imigraccao, lobo2001imigraccao, trento1989outro}. Family and neighbourhood networks created incentives for immigrants to establish trade relations and enforce cooperation through community responsibility systems \citep{roth2014prison}. Because of these particular social characteristics, in the 1890s foreigners were over-represented in the Brazilian trade in general \citep{mattos1991vadios, oliveira2001brasil, truzzi2008patricios} and in the \emph{jogo do bicho} in particular \citep{godoi2012imigraccao, magalhaes2005ganhou, torcato2011repressao, villar2008contravencao}. Although kinship bonds became less relevant over time, these links offered an important element of social cohesion in the \emph{jogo do bicho}'s formative years.

@@

The country's lax financial policy might correlate with poverty through unspecified factors. For instance, political decisions may have caused inadvertently both poverty and the expansion of the monetary base \citep{mattos2013shantytown, schmidt1982modernization}; alternatively, external events such as institutional instability \citep{costantini2014index, fausto2014concise, luna2014economic} or commodity shocks \citep{musacchio2014colonial} could be the cause of those two variables. There is not enough evidence to discard such scenarios. To illustrate this uncertainty, the two nodes, namely, money supply and urban poverty, are connected with a dashed edge in Figure \ref{fig:dag}.

>>>>

In this article I offer a rational choice interpretation of the \emph{jogo do bicho}. More specifically, I discuss how \emph{bicheiros} adopt rational, yet often repressive, strategies to solve collective action problems and maximise their political leverage. I use the case of Rio de Janeiro to illustrate the lottery's profit abilities and its effect in political corruption. In no other part of Brazil the patron-client relationship among \emph{bicheiros}, the civil society, and politicians has been so widely publicised. The \emph{jogo do bicho} entrepreneurs have long been involved in Rio's cultural and financial activities, mainly through the funding of `samba schools' and the Carnival Parade. As I describe below, the parade has played a significant role not only in strenghtening the ties between the \emph{jogo do bicho} and poor urban communities, but also in coopting politicians attracted by the electoral gains offered by the festival \citep{cavalcanti2006carnaval, queiroz1992carnaval}. Furthermore, Brazil's fragmented electoral system creates strong incentives for candidates to use resources from, and ultimately depend on, the lottery's rich sponsors.

@@

\section{Historical Background}
\label{sec:institution}

The late nineteenth-century Brazil had four characteristics that explain how the animal game came to being: 1) a growing urban population excluded from the formal labour market; 2) an inflow of immigrants whose extended family networks helped them engage in trade; 3) an expansion of the monetary supply in the first years of the republic (1880s--1890s); and 4) a judicial system that, albeit repressive, had only imperfect law enforcement.

I start with the impact of urban poverty on the animal game. Brazil abolished slavery in the late 1880s, a period in which the country was rapidly urbanising and freed slaves migrated to its growing cities \citep{andrews1991blacks, skidmore1993black}. The former slaves were joined by increasing numbers of Asian and European immigrants \citep{hall1969origins, lesser2013immigration, smith1979ethnic}. Nevertheless, the job market tightened considerably after the \emph{Encilhamento} financial crisis of 1891 \citep{topik2014political, triner2005baring}. During the economic downturn, the informal economy was the obvious destination for the urban poor. As \citet[115]{chazkel2011laws} observes, one of the few professions open to poor women and foreigners in the early 1900s was that of street vendor. Given the \emph{jogo do bicho} rising popularity, many of those vendors would sell lottery tickets along other types of merchandise.

The immigration also influenced the \emph{jogo do bicho} via social ties. Most foreigners who moved to Brazil came from countries, such as Portugal, Spain or Italy, where extended families were the basic form of social organisation \citep{lobo2001imigraccao, trento1989outro}. Family and neighbourhood networks created incentives for immigrants to establish trade relations and enforce cooperation through community responsibility systems \citep{roth2014prison}. Because of these particular social characteristics, in the 1890s foreigners were over-represented in the Brazilian trade in general \citep{mattos1991vadios, truzzi2008patricios} and in the \emph{jogo do bicho} in particular \citep{godoi2012imigraccao, magalhaes2005ganhou, villar2008contravencao}. Although kinship bonds became less relevant over time, these links offered an important element of social cohesion in the \emph{jogo do bicho}'s formative years.

@@
